The forestecology R package for fitting and assessing neighborhood models of the effect of interspecific competition on the growth of trees

### 2021-08-31 paper/paper_word_output.Rmd
replace all \ref{} with \@ref()

The output `growth_scbi` is a data frame of class `sf` that includes among other variables the species variable `sp` converted to a factor, the average annual `growth` in DBH (cm $\cdot$ y^-1^) for all stems that were alive at both time points, and the `sf` package's encoding of geolocations of `geometry` type `<POINT>`. Given that `growth_scbi` is of class `sf`, it can be easily plotted in `ggplot2` using `geom_sf()` as seen in Figure \ref{fig:scbi-trees}.

@@

Furthermore, we compute two potential competitor explanatory variables $x^{\text{comp}}_{ijk}$ from Equation \ref{eq:model}. First, the basal area of each tree as a function of its DBH in the earlier census. Second, the above ground biomass as estimated by allometric equations encoded in the `get_biomass()` function from the `allodb` package [@allodb_package]; this function has DBH, species, and geographic coordinates as arguments.

@@

Our model of interspecific competition relies on a spatial definition of who competitor trees are: all trees within a distance `comp_dist` of a focal tree. Here we set `comp_dist` to 7.5m, a value informed by other studies [@canham_neighborhood_2004; @uriarte_spatially_2004; @canham_neighborhood_2006], but the package could also be used to compare multiple distances and see which is best supported (see Appendix \ref{mult-dist}). We use `comp_dist` and a manually constructed `sf` representation of the study region's boundary as inputs to `add_buffer_variable()` to add a `buffer` boolean variable to `growth_scbi`. All trees with `buffer` equal to `FALSE` will be our focal trees whose growth will be modeled, whereas those with `TRUE` will only act as competitor trees.

@@

We first manually define an `sf` object defining four folds that partition the study region. We then use the output of the `spatialBlock()` function from the `blockCV` package to associate each tree in `growth_scbi` to the correct `foldID` [@valavi_blockcv_2019]. This `foldID` variable will be used in Section \ref{spatial-cross-validation}.

Figure \ref{fig:scbi-spatial-information} illustrates the net effect of adding these two elements of spatial information to `growth_scbi`.

@@

We then identify all focal trees and their corresponding competitor trees. More specifically, identify all trees that are not part of the buffer region, have a valid `growth` measurement, and have at least one neighbor within `r comp_dist`m. We do this using `create_focal_vs_comp()`, which takes the previously detailed `comp_dist` and `id` arguments, the `sf` representation of the spatial cross-validation blocks `blocks_scbi`, and a specification `comp_x_var` of the `basal_area` variable we use as the competitor explanatory variable $x^{\text{comp}}_{ijk}$ from Equation \ref{eq:model}. This function returns a new data frame `focal_vs_comp_scbi`.

@@

We observe 4 variables describing 20 competitor trees: the unique tree-stem ID, the distance to the focal tree (all $\leq$ `r comp_dist`m), the species, and the basal area (in m$^2$) calculated as $\frac{\pi \times (\text{DBH/2})^2}{10000}$ for the DBH in cm from the earlier census. Saving competitor information in list-column format minimizes redundancy since we do not need to repeat information on the focal tree 20 times. We visualize the spatial distribution of these trees in Figure \ref{fig:scbi-focal-vs-comp-map}.

Here we use basal area as the continuous competitor explanatory variable but the package is flexible to allow the user to specify any competitor explanatory variable (basal area, biomass, tree height, a soil nutrient value, \dots). The package can also be used to compare competitor explanatory variables and see which best explains tree growth, see Appendix \ref{compare-pred-var} for an example comparing basal area and above ground biomass. Similarly, the package can use any categorical variable as an explanatory variable and compare between different categorical variables. For example in @allen_permutation_2020 we compare grouping individuals based on species, family, and based on trait-based groups. In Appendix \ref{compare-group-var} we give another example and compare grouping individuals by species or by potential canopy position (canopy, understory, shrub layer).

@@

Lastly, we fit the competition Bayesian linear regression model for tree growth outlined in Equation \ref{eq:model} using `comp_bayes_lm()`. This function has an option to specify prior distributions of all parameters, chosen here to be the defaults detailed in `?comp_bayes_lm`.

@@

Lastly, the generic for `ggplot2::autoplot()` allows us to visualize all posterior distributions, as seen in Figure \ref{fig:scbi-posterior-viz}. Setting `type` to `"intercepts"` and `"dbh_slopes"` returns species-specific posterior distributions for $\beta_{0,j}$ and $\beta_{dbh,j}$ respectively, while setting `type = "competition"` returns competition coefficients $\lambda_{j,k}$.

@@

Currently the `forestecology` package can only fit the competition Bayesian linear regression model in Equation \ref{eq:model}. However, it can be extended to any model as long as it is implemented in a function similar to `comp_bayes_lm()`.

@@

To evaluate the effect of competitor species identity, we use the above four steps along with the permutation test in Equation \ref{eq:permutation-hypothesis-test}. Under a null hypothesis where competitor species identity does not matter, we can permute the competitor species identities within each focal tree, compute the RMSE test statistic, repeat this process several times to construct a null distribution, and compare it to the observed RMSE to assess significance. Going back to our example in Section \ref{focal-vs-comp} of focal tree with `focal_ID` 4 and its 20 competitors, the permutation test only randomly resamples the `comp_sp` variable without replacement, leaving all other variables intact. This resampling is nested within each focal tree in order to preserve neighborhood structure. We perform this permutation test once again using `comp_bayes_lm()` but by setting `run_shuffle = TRUE`.

@@

To evaluate model performance, we use spatial cross-validation. The model fit in Section \ref{model-fit-predict} uses the same data to both fit and assess model performance. Given the spatial-autocorrelation of our data, this can potentially lead to overfit models [@roberts_cross-validation_2017]. To mitigate this risk, we use the spatial cross-validation blocking scheme encoded in the `foldID` variable from Section \ref{spatial-information} and visualized in Figure \ref{fig:scbi-spatial-information}.

At each iteration of the cross-validation, one fold acts as the test set and the remaining three act as the training set. We fit the model to all focal trees in the training set, apply the model to all focal trees in the test set, compute predicted values, and compute the RMSE. Furthermore, to maintain spatial independence between the test and training sets, a "fold buffer" that extends `r comp_dist`m outwards from the boundary of the test set is considered; all trees within this "fold buffer" are excluded from the training set (see Figure \ref{fig:scbi-spatial-cross-validation-schematic}).

@@

`run_cv()` also accepts the `run_shuffle` argument in order to permute competitor species identity as described in Section \ref{permutation-test}. Figure \ref{fig:scbi-simulation} compares model performance for 49 permutations of competitor species and RMSE calculations, both with and without cross-validation. Without cross-validation, competitor species identity does matter as the observed RMSE was significantly lower than the permutation null distribution of RMSE. However, once we incorporate spatial cross-validation, this improvement disappears. These results suggest that in this 9 ha subplot of the SCBI plot, competitive interactions do not depend on the identity of the competitor, which is the opposite of what has been observed in other locations [@allen_permutation_2020; @uriarte_spatially_2004]. This provides a striking example of the importance of cross-validation, as without it the over-fit model gives rise to an incorrect conclusion.

@@

We also hope to extend the `forestecology` package's functionality to account for a larger variety of models for tree growth. One clear future direction would be to allow competition based on species trait values rather than species identity. There is evidence that traits predict competitive outcomes [@kunstler_competitive_2012; @lasky_trait-mediated_2014; @uriarte_trait_2010]. Thus an extension of the model would allow $\lambda$ values from Equation \ref{eq:model} to be a function of the traits of competing species.

Lastly, the `forestecology` current uses the `blockCV` package behind the scenes to create the spatial blocks acting as folds for our spatial cross-validation algorithm detailed in Sections \ref{spatial-information} and \ref{spatial-cross-validation} [@valavi_blockcv_2019]. This back-end functionality could be substituted with the `spatialsample` package for spatial resampling infrastructure; a `tidymodels` package under active development as of 2021 [@spatialsample_package; @tidymodels_package].

Lastly, currently the package only implements the Bayesian linear regression model detailed in Equation \ref{eq:model}. As we demonstrate in Section \ref{model-fit-predict} however, the fitting of this model is self-contained in a single function `comp_bayes_lm()` which returns an object of S3 class type `comp_bayes_lm`. This class has generic methods implemented to print, make predictions, and plot all results. Therefore the package can be modularly extended to fit other models as long as they are coded similarly to `comp_bayes_lm()` and have equivalent generic methods implemented.

@@

We observe in Figure \ref{fig:appendix2-plot} that a cut-off distance of approximately 6m minimizes the cross-validation estimated RMSE.

@@

In the above code we use the basal area of an individual as a continuous competitor explanatory variable. But the package allows the user to specify any competitor explanatory variable in the `comp_x_var` argument of `create_focal_vs_comp` function. Here we use the cross-validated model comparison to see which of two possible competitor explanatory variables computed in Section \ref{compute-growth}, basal area or above ground biomass, best explains growth.

@@

Here we observe that basal area is a better competitor explanatory variable competitor explanatory variables $x^{\text{comp}}_{ijk}$ from Equation \ref{eq:model} than above ground biomass as suggested by the lower estimated RMSE.

@@

We find that species identity has a lower RMSE, so does a better job. We still however plot the competition posteriors for the canopy position groupings in Figure \ref{fig:can-pos-comp-plot}. Unsurprisingly we see that canopy and canopy emergent competitors generally have negative effects on their neighbors, while shrubs and understory competitors have neutral or even positive effects.

@@

This code replicates Figure \ref{fig:scbi-simulation}: A comparison of root mean squared error of
models for standard, permuted, and spatial cross-validated error estimates.

>>>>

The output `growth_scbi` is a data frame of class `sf` that includes among other variables the species variable `sp` converted to a factor, the average annual `growth` in DBH (cm $\cdot$ y^-1^) for all stems that were alive at both time points, and the `sf` package's encoding of geolocations of `geometry` type `<POINT>`. Given that `growth_scbi` is of class `sf`, it can be easily plotted in `ggplot2` using `geom_sf()` as seen in Figure \@ref(fig:scbi-trees).

@@

Furthermore, we compute two potential competitor explanatory variables $x^{\text{comp}}_{ijk}$ from Equation \@ref(eq:model). First, the basal area of each tree as a function of its DBH in the earlier census. Second, the above ground biomass as estimated by allometric equations encoded in the `get_biomass()` function from the `allodb` package [@allodb_package]; this function has DBH, species, and geographic coordinates as arguments.

@@

Our model of interspecific competition relies on a spatial definition of who competitor trees are: all trees within a distance `comp_dist` of a focal tree. Here we set `comp_dist` to 7.5m, a value informed by other studies [@canham_neighborhood_2004; @uriarte_spatially_2004; @canham_neighborhood_2006], but the package could also be used to compare multiple distances and see which is best supported (see Appendix \@ref(mult-dist)). We use `comp_dist` and a manually constructed `sf` representation of the study region's boundary as inputs to `add_buffer_variable()` to add a `buffer` boolean variable to `growth_scbi`. All trees with `buffer` equal to `FALSE` will be our focal trees whose growth will be modeled, whereas those with `TRUE` will only act as competitor trees.

@@

We first manually define an `sf` object defining four folds that partition the study region. We then use the output of the `spatialBlock()` function from the `blockCV` package to associate each tree in `growth_scbi` to the correct `foldID` [@valavi_blockcv_2019]. This `foldID` variable will be used in Section \@ref(spatial-cross-validation).

Figure \@ref(fig:scbi-spatial-information) illustrates the net effect of adding these two elements of spatial information to `growth_scbi`.

@@

We then identify all focal trees and their corresponding competitor trees. More specifically, identify all trees that are not part of the buffer region, have a valid `growth` measurement, and have at least one neighbor within `r comp_dist`m. We do this using `create_focal_vs_comp()`, which takes the previously detailed `comp_dist` and `id` arguments, the `sf` representation of the spatial cross-validation blocks `blocks_scbi`, and a specification `comp_x_var` of the `basal_area` variable we use as the competitor explanatory variable $x^{\text{comp}}_{ijk}$ from Equation \@ref(eq:model). This function returns a new data frame `focal_vs_comp_scbi`.

@@

We observe 4 variables describing 20 competitor trees: the unique tree-stem ID, the distance to the focal tree (all $\leq$ `r comp_dist`m), the species, and the basal area (in m$^2$) calculated as $\frac{\pi \times (\text{DBH/2})^2}{10000}$ for the DBH in cm from the earlier census. Saving competitor information in list-column format minimizes redundancy since we do not need to repeat information on the focal tree 20 times. We visualize the spatial distribution of these trees in Figure \@ref(fig:scbi-focal-vs-comp-map).

Here we use basal area as the continuous competitor explanatory variable but the package is flexible to allow the user to specify any competitor explanatory variable (basal area, biomass, tree height, a soil nutrient value, \dots). The package can also be used to compare competitor explanatory variables and see which best explains tree growth, see Appendix \@ref(compare-pred-var) for an example comparing basal area and above ground biomass. Similarly, the package can use any categorical variable as an explanatory variable and compare between different categorical variables. For example in @allen_permutation_2020 we compare grouping individuals based on species, family, and based on trait-based groups. In Appendix \@ref(compare-group-var) we give another example and compare grouping individuals by species or by potential canopy position (canopy, understory, shrub layer).

@@

Lastly, we fit the competition Bayesian linear regression model for tree growth outlined in Equation \@ref(eq:model) using `comp_bayes_lm()`. This function has an option to specify prior distributions of all parameters, chosen here to be the defaults detailed in `?comp_bayes_lm`.

@@

Lastly, the generic for `ggplot2::autoplot()` allows us to visualize all posterior distributions, as seen in Figure \@ref(fig:scbi-posterior-viz). Setting `type` to `"intercepts"` and `"dbh_slopes"` returns species-specific posterior distributions for $\beta_{0,j}$ and $\beta_{dbh,j}$ respectively, while setting `type = "competition"` returns competition coefficients $\lambda_{j,k}$.

@@

Currently the `forestecology` package can only fit the competition Bayesian linear regression model in Equation \@ref(eq:model). However, it can be extended to any model as long as it is implemented in a function similar to `comp_bayes_lm()`.

@@

To evaluate the effect of competitor species identity, we use the above four steps along with the permutation test in Equation \@ref(eq:permutation-hypothesis-test). Under a null hypothesis where competitor species identity does not matter, we can permute the competitor species identities within each focal tree, compute the RMSE test statistic, repeat this process several times to construct a null distribution, and compare it to the observed RMSE to assess significance. Going back to our example in Section \@ref(focal-vs-comp) of focal tree with `focal_ID` 4 and its 20 competitors, the permutation test only randomly resamples the `comp_sp` variable without replacement, leaving all other variables intact. This resampling is nested within each focal tree in order to preserve neighborhood structure. We perform this permutation test once again using `comp_bayes_lm()` but by setting `run_shuffle = TRUE`.

@@

To evaluate model performance, we use spatial cross-validation. The model fit in Section \@ref(model-fit-predict) uses the same data to both fit and assess model performance. Given the spatial-autocorrelation of our data, this can potentially lead to overfit models [@roberts_cross-validation_2017]. To mitigate this risk, we use the spatial cross-validation blocking scheme encoded in the `foldID` variable from Section \@ref(spatial-information) and visualized in Figure \@ref(fig:scbi-spatial-information).

At each iteration of the cross-validation, one fold acts as the test set and the remaining three act as the training set. We fit the model to all focal trees in the training set, apply the model to all focal trees in the test set, compute predicted values, and compute the RMSE. Furthermore, to maintain spatial independence between the test and training sets, a "fold buffer" that extends `r comp_dist`m outwards from the boundary of the test set is considered; all trees within this "fold buffer" are excluded from the training set (see Figure \@ref(fig:scbi-spatial-cross-validation-schematic).

@@

`run_cv()` also accepts the `run_shuffle` argument in order to permute competitor species identity as described in Section \@ref(permutation-test). Figure \@ref(fig:scbi-simulation) compares model performance for 49 permutations of competitor species and RMSE calculations, both with and without cross-validation. Without cross-validation, competitor species identity does matter as the observed RMSE was significantly lower than the permutation null distribution of RMSE. However, once we incorporate spatial cross-validation, this improvement disappears. These results suggest that in this 9 ha subplot of the SCBI plot, competitive interactions do not depend on the identity of the competitor, which is the opposite of what has been observed in other locations [@allen_permutation_2020; @uriarte_spatially_2004]. This provides a striking example of the importance of cross-validation, as without it the over-fit model gives rise to an incorrect conclusion.

@@

We also hope to extend the `forestecology` package's functionality to account for a larger variety of models for tree growth. One clear future direction would be to allow competition based on species trait values rather than species identity. There is evidence that traits predict competitive outcomes [@kunstler_competitive_2012; @lasky_trait-mediated_2014; @uriarte_trait_2010]. Thus an extension of the model would allow $\lambda$ values from Equation \@ref(eq:model) to be a function of the traits of competing species.

Lastly, the `forestecology` current uses the `blockCV` package behind the scenes to create the spatial blocks acting as folds for our spatial cross-validation algorithm detailed in Sections \@ref(spatial-information) and \@ref(spatial-cross-validation) [@valavi_blockcv_2019]. This back-end functionality could be substituted with the `spatialsample` package for spatial resampling infrastructure; a `tidymodels` package under active development as of 2021 [@spatialsample_package; @tidymodels_package].

Lastly, currently the package only implements the Bayesian linear regression model detailed in Equation \@ref(eq:model). As we demonstrate in Section \@ref(model-fit-predict) however, the fitting of this model is self-contained in a single function `comp_bayes_lm()` which returns an object of S3 class type `comp_bayes_lm`. This class has generic methods implemented to print, make predictions, and plot all results. Therefore the package can be modularly extended to fit other models as long as they are coded similarly to `comp_bayes_lm()` and have equivalent generic methods implemented.

@@

We observe in Figure \@ref(fig:appendix2-plot) that a cut-off distance of approximately 6m minimizes the cross-validation estimated RMSE.

@@

In the above code we use the basal area of an individual as a continuous competitor explanatory variable. But the package allows the user to specify any competitor explanatory variable in the `comp_x_var` argument of `create_focal_vs_comp` function. Here we use the cross-validated model comparison to see which of two possible competitor explanatory variables computed in Section \@ref(compute-growth), basal area or above ground biomass, best explains growth.

@@

Here we observe that basal area is a better competitor explanatory variable competitor explanatory variables $x^{\text{comp}}_{ijk}$ from Equation \@ref(eq:model) than above ground biomass as suggested by the lower estimated RMSE.

@@

We find that species identity has a lower RMSE, so does a better job. We still however plot the competition posteriors for the canopy position groupings in Figure \@ref(fig:can-pos-comp-plot). Unsurprisingly we see that canopy and canopy emergent competitors generally have negative effects on their neighbors, while shrubs and understory competitors have neutral or even positive effects.

@@

This code replicates Figure \@ref(fig:scbi-simulation): A comparison of root mean squared error of
models for standard, permuted, and spatial cross-validated error estimates.
